---
title: "team.Rmd"
author: "Dana Thacker, Nikole Slinger"
date: "4/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", 
                stringsAsFactors = FALSE)
per <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", 
                stringsAsFactors = FALSE)
```

**Question 1**
Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?
```{r}

```

**Question 2**
Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)
```{r}

```

**Question 3**
Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)
```{r}

```

**Question 4**
Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)
```{r}

```

**Question 5**
Join the accident and person table (work out which variable(s) to use)
```{r}

```

**Question 6**
Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results and explain what you find.
```{r}

```

**Question 7**
Now plot a choropleth map of the number of deaths on a county level. Also explain what you find.
```{r}

```

**Question 8**
Is summer or winter more dangerous? Does this depend on states? Explore and explain.
```{r}

```

